Computing Exact Restricted Isometry Constants via\\ Mixed Integer Semidefinite Programming

### 2016-03-12 RIP-MISDP-Paper/IEEE_RIP-MISDP.tex
rework MISDP section

\section{An MISDP for the Restricted Isometry Property}
\label{sec:MISDP}

In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{lowerRIP}) and (\ref{upperRIP}) using mixed integer semidefinite programming. We will again start with the non-convex quadratic problem
\begin{align}\label{QP2}
 \min \quad & \Norm{Ax}{2}^2 \nonumber \\
 \text{s.t.} \quad & \Norm{x}{2}^2 = 1 \tag{QP} \\
 & \Norm{x}{0} \leq k. \nonumber
\end{align}
For the non-convex quadratic equality constraint $\Norm{x}{2}^2 = 1$ we will
again use semidefinite lifting, introducing the new variable $X=xx\T$ like in the last section. 
The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. We can thus rewrite (\ref{QP}) equivalently as
% In this section we want to compute the optimal constants $\alpha_k$ and $\beta_k$ in (\ref{asymRIP}) using mixed integer semidefinite programming. A non-convex quadratic optimization problem to compute $\alpha_k$ can be formulated as
% 
% \begin{align}\label{QP}
%  \min \quad & \Norm{Ax}{2}^2 \nonumber \\
%  \text{s.t.} \quad & \Norm{x}{2}^2 = 1 \tag{QP} \\
%  & \Norm{x}{0} \leq k. \nonumber
% \end{align}
% 
% By switching $\min$ for $\max$ in (\ref{QP}) we also get a corresponding optimization problem to compute $\beta_k$ and therefore also $\gamma_k$ and $\delta_k$.
% The cardinality-constraint in (\ref{QP}) can equivalently be written using binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i \leq k$. For the non-convex quadratic equality constraint $\Norm{x}{2}^2 = 1$ we will
% use the technique of semidefinite lifting first introduced in \cite{GW95} for the max-cut problem. The idea is to use a new matrix variable $X=xx\T$ for the quadratic terms, where the condition $X=xx\T$ is then enforced by the 
% equivalent constraints $X \succeq 0$ and $\Rk(X) = 1$, where $X
% \succeq 0$ means that $X$ is symmetric and positive semidefinite. Using
% that $\Norm{Ax}{2}^2 = x\T A\T Ax = \Tr(A\T A X)$ and 
% $\Tr(X) = \sum_{i=1}^n x_i^2$, we can thus rewrite (\ref{QP}) equivalently to
\begin{align}\label{Rk1MISDP}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \text{for } j = 1, ..., n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{M1} \\
 & \Rk(X) = 1 \nonumber \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n. \nonumber
\end{align}
The next step would then be to relax the non-convex rank constraint again to arrive at the following relaxation which computes a lower bound on the optimal constant $\alpha_k$ in (\ref{lowerRIP}):
\begin{align}\label{MISDP}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \text{for } j = 1, ..., n \nonumber \\
 & \sum_{i=1}^n z_i \leq k \tag{MISDP} \\
 & X \succeq 0  \nonumber \\
 & z \in \{0,1\}^n \nonumber
\end{align}

In the following we will show that (\ref{MISDP}) not only computes a lower bound but the exact value of $\alpha_k$. For this we will use results about the geometry of semidefinite programs to show that there always exists an 
optimal solution to (\ref{MISDP}) with rank 1 so that the optimal objective values of (\ref{Rk1MISDP}) and (\ref{MISDP}) have to agree. We will use the following result by Pataki \cite{pat98} about the relationship between the rank of
solutions of semidefinite programs and the dimension of corresponding faces, meaning a subset $F$ of the convex set $P$ such that \mbox{$x \in F, y, z \in P$} and $x = 0.5y + 0.5z$ imply $y,z \in F$:

\begin{theorem}\label{patakiLemma}
 Let $X \in F$, where $F$ is a face of
 \begin{equation*}
 P \define \{X \in S_n : X \succeq 0, \Tr(A_i X) = b_i \ \text{for } i = 1, ..., m\}
 \end{equation*}
 for symmetric matrices $A_i \in S_n, i \leq m$. Then
 \begin{equation*}
  \frac{1}{2}\Rk(X)\cdot(\Rk(X)+1) \leq m + dim(F).
 \end{equation*}
\end{theorem}

Now we will use this theorem on the projection of the feasible set of (\ref{MISDP}) onto the space of all indices $i$, such that $z_i = 1$ for a given optimal solution of (\ref{MISDP}), to show that there always exists an optimal 
solution to (\ref{MISDP}) with rank one.

\begin{theorem}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\Rk(X^*) = 1$. 
\end{theorem}
\begin{proof}
 Because of the constraint $\Tr(X) = 1$, Problem~\eqref{MISDP} is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1,$ \mbox{$z_1 = 1, z_i = 0 \ \forall i \neq 1$} 
 is always feasible for $k > 0$,  so an optimal solution to~\eqref{MISDP} exists.
 Let $\hat{X}, \hat{z}$ be such an optimal solution to (\ref{MISDP}) with $\sum_{i=1}^n z_i =: \ell \leq k$.
 Define
 \begin{align}\label{Proj}
 \min \quad & \Tr(\tilde{A}\T\tilde{A}\tilde{X}) \nonumber \\
 \text{s.t.} \quad & \Tr(\tilde{X}) = 1 \tag{Proj} \\
 & \tilde{X} \succeq 0  \nonumber 
 \end{align}
for $\tilde{X} \in S_\ell$, where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those
columns $\{{i_1}, {i_2}, ..., {i_\ell}\} \enifed T$ of $A$ for which
$\hat{z}_{i_k} = 1$.

Now let $\bar{X}$ be the $\ell \times \ell$ submatrix of $\hat{X}$ with all rows and columns outside of $T$ removed. Then $\bar{X}$ is feasible for (\ref{Proj}) with the same objective value as the optimal solution $\hat{X}$ of (\ref{MISDP}), 
as we only removed zero rows and columns of $\hat{X}$ and $(A\T A)\hat{X}$, which do not influence the trace or the positive semidefiniteness.

On the other hand we can lift any optimal solution $\check{X}$ of (\ref{Proj}) to a feasible point for (\ref{MISDP}) with identical objective value by setting $z^{\prime}_i$ to one if and only if $i \in T$
and $X^{\prime}_{i_{j_1} i_{j_2}} = \check{X}_{j_1j_2}$ if and only if $i_{j_1}, i_{j_2} \in T$
and $X^{\prime}_{ij}=0$ otherwise. This solution will then be feasible for (\ref{MISDP}), since $\Tr(\check{X}) = 1$ and $\check{X}\in S_\ell$ imply $0 \leq \check{X}_{jj} = X^{\prime}_{i_j i_j} \leq 1$ and therefore 
$X^{\prime}_{ij} \leq z^{\prime}_j$ by diagonal dominance of semidefinite matrices.

So the optimal objective values of (\ref{Proj}) and (\ref{MISDP}) agree and with this construction also 
$\Rk(X^{\prime}) = \Rk(\check{X})$, so it suffices to show that (\ref{Proj}) always allows for an optimal solution of rank one.
Let $\breve{X}$ be an extreme point of (\ref{Proj}) (as the set of optimal points of (\ref{Proj}) is a non-empty, convex, compact subset of $\R^{\ell(\ell+1)/2}$, such a point exists
\cite[Corollary 18.5.1]{Roc70}) with corresponding face $F=\{\breve{X}\}$. Then using Theorem~\ref{patakiLemma} on (\ref{Proj}) with $\tilde{m} = 1$ LP-constraint gives
\begin{equation*}
 \frac{1}{2}\Rk(\breve{X})\cdot(\Rk(\breve{X})+1) \leq \tilde{m} + dim(F) = 1 + 0 = 1.
\end{equation*}
Since $\breve{X}=0$ is infeasible for $\Tr(X)=1$, we must have $\Rk(\breve{X}) = 1$, which can then be lifted to an optimal solution of (\ref{MISDP}) with rank one.
\end{proof}

Since Theorem~\ref{Rk1thm} guarantees that (\ref{MISDP}) always has a solution of rank one and the rank-constraint was the only constraint we relaxed to get (\ref{MISDP}) from (\ref{QP}), this means that their optimal objective values 
have to agree and we can compute the optimal constant $\alpha_k$ for the RIP by solving (\ref{MISDP}). Since the same argumentation also holds for the maximization problem for $\beta_k$, this allows us to compute the restricted isometry
constant and the restricted isometry ratio by mixed integer semidefinite programming.

\begin{remark}\label{Rk1rmk}
Theorem \ref{Rk1thm} still holds if additional constraints are added fixing some of the binary variables $z_j$ to $0$ or $1$. In this case there will always exist an optimal solution to problem (\ref{MISDP}) with these added
constraints with rank $1$, such that the optimal objective value of this problem will also equal that of (\ref{Rk1MISDP}) with the added constraints. This means in particular, that solving problem (\ref{MISDP}) using a 
branch-and-bound-approach is the same as solving the original problem (\ref{Rk1MISDP}) with a branch-and-bound-code.
\end{remark}

To solve \eqref{MISDP} successfully using a branch-and-bound-approach, it is important to have strong semidefinite relaxations for fractional $z$ variables. 
Therefore we will further strengthen the constraint coupling the continuous and binary variables to force the nonzero entries of $z$ towards one in the
semidefinite relaxations.

\begin{lemma}\label{HalfConstraint} 
The constraint $-z_j \leq X_{ij} \leq z_j$ in~\eqref{MISDP} can be substituted for all non-diagonal entries of $X$ by the stronger constraint 
\begin{equation}\label{eq:HalfConstraint}
  -\frac{1}{2}z_j \leq X_{ij} \leq \frac{1}{2}z_j.
\end{equation}
\end{lemma}
\begin{proof}
Let $X$ satisfy $\Tr(X)=1$ and $X\succeq 0$. Then $X_{ii}+X_{jj}\leq 1$ and nonnegativity of the diagonal 
entries imply $X_{ii}X_{jj}\leq 1/4$. 
Since all minors of $X$ need to be nonnegative, it follows that
\begin{equation*}
  X_{ij}^2 \leq  X_{ii} X_{jj} \leq \frac{1}{4}
\end{equation*}
and therefore $-1/2 \leq X_{ij} \leq 1/2$, which implies the equivalence between
$-z_j \leq X_{ij} \leq z_j$ and~\eqref{eq:HalfConstraint} for all integer values 
of $z$.
\end{proof}

Theorem \ref{Rk1thm} and Remark \ref{Rk1rmk} do not guarantee that any found solution will have rank one, they only guarantee that there exists a rank one solution. But this is not a problem for applications in the context of the RIP, 
since we are only interested in the optimal objective value, since it implies the constant $\alpha_k$ or $\beta_k$ for the RIP. For other applications like the SPCA such an extreme solution can also be generated with an algorithm 
proposed by Pataki in \cite{coneLP}.

>>>>

\section{An MISDP for the RIP}
\label{sec:MISDP}

\noindent
In this section we propose a MISDP formulation to compute the optimal
constants $\alpha_k$ and $\beta_k$ in \eqref{lowerRIP} and
\eqref{upperRIP}, respectively. We will again start with the non-convex quadratic problem~\eqref{QP}.
For the non-convex quadratic equality constraint $\Norm{x}{2}^2 = 1$, we
again use semidefinite lifting, introducing the new variable $X=xx\T$ like in the last section. 
The cardinality-constraint in \eqref{QP} can equivalently be written using
binary variables $z_i$ as $-z_i \leq x_i \leq z_i$ and $\sum_{i=1}^n z_i
\leq k$. We can thus equivalently rewrite \eqref{QP} as
\begin{align}\label{Rk1MISDP}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1, \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \text{for } j = 1, \dots, n, \nonumber \\
 & \sum_{i=1}^n z_i \leq k,\\
 & \Rk(X) = 1, \nonumber \\
 & X \succeq 0,  \nonumber \\
 & z \in \{0,1\}^n. \nonumber
\end{align}
We then relax the non-convex rank constraint to arrive at the following
MISDP formulation:
\begin{align}\label{MISDP}
 \min \quad & \Tr(A\T A X) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1, \nonumber \\
 & -z_j \leq X_{ij} \leq z_j \quad \text{for } j = 1, \dots, n, \nonumber \\
 & \sum_{i=1}^n z_i \leq k, \tag{MISDP} \\
 & X \succeq 0,  \nonumber \\
 & z \in \{0,1\}^n. \nonumber
\end{align}
In the following, we will show that \eqref{MISDP} actually produces the
exact value of $\alpha_k$. We will use the following result about the
relationship between the rank of solutions of semidefinite programs and the
dimension of corresponding faces of a convex set~$P$, i.e., subsets $F
\subseteq P$ such that $x \in F$, $y, z \in P$, and $x = \tfrac{1}{2} (y + z)$ imply
$y,z \in F$:

\begin{theorem}[Pataki~\cite{pat98}]\label{patakiLemma}
  Let $X \in F$, where $F$ is a face of
  \begin{equation*}
    P \define \{X \in S_n \suchthat X \succeq 0,\; \Tr(A_i X) = b_i\; \text{ for } i = 1, \dots, \tilde{m}\}
  \end{equation*}
  for symmetric matrices $A_i \in S_n$, $i = 1, \dots, \tilde{m}$. Then
  \begin{equation*}
    \tfrac{1}{2}\Rk(X)\cdot(\Rk(X)+1) \leq \tilde{m} + dim(F).
  \end{equation*}
\end{theorem}

Using this result on the projection of~\eqref{MISDP} onto the $X$ variables, yields the following.

\begin{theorem}\label{Rk1thm}
  For every $A \in \R^{m \times n}$ and $k \in \Z_{>0}$, there exists an optimal solution $(X^\star, z^\star)$ of \eqref{MISDP} with $\Rk(X^\star) = 1$. 
\end{theorem}

\begin{proof}
  First note that because of the constraint $\Tr(X) = 1$,
  Problem~\eqref{MISDP} is bounded. Moreover, $X_{11} = 1$, $X_{ij} = 0$
  for all $i,j \neq 1$, $z_1 = 1$, $z_i = 0$ for all $i = 2, \dots, n$ is
  feasible for $k > 0$. Thus, an optimal solution to~\eqref{MISDP} exists.

  Let $(X^\star, z^\star)$ be an optimal solution to \eqref{MISDP} with
  $\ell \define \sum_{i=1}^n z^\star_i \leq k$. Let $T \define \{i_1, i_2,
  \dots, i_\ell\}$ be the support of $z^\star$, and define $\check{A} =
  (A_{i_1}, A_{i_2}, \dots, A_{i_\ell})$ to be the submatrix of~$A$ formed by
  the columns indexed by~$T$. Then consider
  \begin{align}\label{Proj}
    \min \quad & \Tr(\check{A}\T \check{A} \check{X}), \nonumber \\
    \text{s.t.} \quad & \Tr(\check{X}) = 1, \\
    & \check{X} \succeq 0,\; \check{X} \in S_\ell.\nonumber 
  \end{align}

  Let $\check{X}$ be the $\ell \times \ell$ submatrix of $X^\star$ with all
  rows and columns outside of $T$ removed. Then $\check{X}$ is feasible for
  \eqref{Proj} with the same objective value, since we only removed zero
  rows and columns of $X^\star$ and $(A\T A) X^\star$, which do not
  influence the trace or the positive semidefiniteness.

  On the other hand, we can lift any solution $\check{X}$ of~\eqref{Proj}
  to a point $\hat{X} \in S_n$ by extending with zeros. Defining
  $\hat{z}_i$ to be 1 if and only if $i \in T$, yields a feasible point
  $(\hat{X}, \hat{z})$ for \eqref{MISDP} with identical objective value,
  since $\Tr(\check{X}) = 1$ and $\check{X} \in S_\ell$ imply $0 \leq
  \check{X}_{jj} = \hat{X}_{i_j i_j} \leq 1$ for $j = 1, \dots,
  \ell$. Therefore, $\hat{X}_{ij} \leq \hat{z}_j$ by diagonal dominance of
  semidefinite matrices.

  Thus, the optimal objective values of \eqref{Proj} and \eqref{MISDP}
  agree and $\Rk(\hat{X}) = \Rk(\check{X})$.  Therefore, it
  suffices to show that \eqref{Proj} has an optimal solution of rank one.

  Let $\check{X}$ be an extreme point of \eqref{Proj} (since the set of
  optimal points of~\eqref{Proj} is a nonempty, convex, and compact, such a
  point exists, see~\cite[Corollary 18.5.1]{Roc70}). Defining the face $F =
  \{\check{X}\}$ and applying Theorem~\ref{patakiLemma} on \eqref{Proj}
  gives
  \begin{equation*}
    \tfrac{1}{2} \Rk(\check{X}) \cdot (\Rk(\check{X})+1) \leq \tilde{m} + dim(F) = 1 + 0 = 1,
  \end{equation*}
  since there are $\tilde{m} = 1$ linear constraints. Since $\check{X} = 0$
  is infeasible for $\Tr(X) = 1$, it follows that $\Rk(\check{X}) = 1$,
  which can then be lifted to an optimal solution of \eqref{MISDP} of rank
  one.
\end{proof}

Theorem~\ref{Rk1thm} guarantees that \eqref{MISDP} always has a solution of
rank one. Therefore, the optimal values of \eqref{MISDP} and \eqref{QP}
agree. We can therefore compute the optimal constant $\alpha_k$ for the RIP
by solving \eqref{MISDP}. Since the same argumentation also holds for the
maximization problem for $\beta_k$, this allows us to compute the
restricted isometry constant and the restricted isometry ratio by
mixed-integer semidefinite programming.

\begin{remark}\label{Rk1rmk}
  Theorem \ref{Rk1thm} still holds, if additional constraints involving
  only $z$-variables are added, e.g., fixing some $z_j$ to $0$ or $1$. In
  particular, \eqref{MISDP} or equivalently \eqref{Rk1MISDP} can be
  solved by branch and bound.
\end{remark}

Theorem~\ref{Rk1thm} and Remark~\ref{Rk1rmk} do not guarantee that any
found solution will have rank one. However, branch and bound will still
produce the correct optimal value. This suffices for our application of
computing the RIP constants $\alpha_k$ and $\beta_k$. For other
applications like SPCA, an extreme solution can be generated with an
algorithm proposed by Pataki~\cite{coneLP}.

In order to successfully apply branch and bound to \eqref{MISDP}, it is
important to use strong semidefinite relaxations for fractional $z$
variables. To this end, we can use the following strengthening of the
constraints $-z_j \leq X_{ij} \leq z_j$.

\begin{lemma}\label{HalfConstraint} 
  The inequalities
  \begin{equation}\label{eq:HalfConstraint}
    -\tfrac{1}{2} z_j \leq X_{ij} \leq \tfrac{1}{2} z_j,
  \end{equation}
  for $i, j \in \{1, \dots, n\}$, are valid for~\eqref{MISDP}.
\end{lemma}

\begin{proof}
  Let $X \succeq 0$ satisfy $\Tr(X)=1$. Then $X_{ii}+X_{jj}\leq 1$ and
  nonnegativity of the diagonal entries imply $X_{ii}X_{jj}\leq
  \tfrac{1}{4}$.  Since all minors of $X$ need to be nonnegative, it
  follows that
  \begin{equation*}
    X_{ij}^2 \leq  X_{ii} X_{jj} \leq \tfrac{1}{4}.
  \end{equation*}
  Therefore $- \tfrac{1}{2} \leq X_{ij} \leq \tfrac{1}{2}$, which shows
  validity of~\eqref{eq:HalfConstraint}.
\end{proof}
